Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from the ERA5-Land and MODIS

### 2023-10-09 article/paper_drought_Chile2023.qmd
first draft of the introduction

title: Drought assessment of the impact of water supply and demand over land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS datasets

@@

@@

Precipitation is the primary driver of drought, which impacts soil moisture, hydrological regimes, and vegetation productivity. Initially, drought was commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans' decisions and activities that drives the anthropogenic drought. Even though it has been argued that those definitions do not fully address the ecological dimensions of drought. @Crausbay2017 proposed the ecological drought definition as *"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"*. Moreover, many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019]. On the other hand, the AR6 [@IPCC2023] states that even if global warming is stabilized at 1.5°–2°C, many parts of the world will be impacted by more severe agricultural and ecological droughts. Then, there is a challenge in conducting drought research, especially its impact on ecosystems.

Chile has been facing a persistent rainfall deficit for more than a decade [@Garreaud2017], which has impacted the the vegetation development [@Zambrano2023] and hydrological system [@Boisier2018]. Current drought conditions have affected crop yields in Central Chile. Highlighting the growing seasons 2007-2008 and 2008-2009 [@Zambrano2016, @Zambrano2018], which impacted an extensive surface in [COMPLETAR]. But, in 2019–2020, the drought intensity reached an extreme condition in North 34°S not seen for at least 40 years [@Zambrano2023], thus affecting forest, grassland, and cropland areas. The prolonged lack of precipitation in Central Chile is producing changes in ecosystem dynamics that must be studied.

Satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how meteorological drought impacts vegetation dynamics. Since the 90’s multiple vegetation drought indices have been developed, such as (VCI,@Kogan1990; TCI, @Kogan1995; zNDVI, [@Peters2002]; VegDri, [@Brown2008]) that have allowed making spatiotemporal analysis. Although we can derive these indices for any time during the year (depending on satellite revisit), they are more relevant during the the growing season [@Mishra2015]. Although modeling phenology is a complex task (e.g, ….), satellites offer strategies that help to address these issues [@Younes2021; @Vrieling2018; @Cai2017]. Moreover, land cover dynamics products such as the MCD12Q2 from the USGS [@Friedl2019] provide phenology metrics. Rather than estimating these parameters at a given point in time, some authors have proposed aggregating these indices throughout the season. @Meroni2017 proposes to accumulate the fractional active photosynthetic active radiation (FAPAR) between the start (SOS), and the end of the season (EOS) thus obtaining zCFAPAR as a proxy for biomass productivity. @Zambrano2018 used the same approach but with the NDVI (Normalized Difference Vegetation Index), deriving the zcNDVI for Central Chile. Besides, land use land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021]. To analyze these changes, multiple time-series LULC products exist, such as the MCD12Q1 [@Friedl2019] and the ESA CCI-LC (ESA 2017). The LULC product, together with the vegetation drought index can help evaluate the impact of drought on the ecosystem.

Vegetation drought indices are commonly used as proxies of productivity [@Paruelo2016; @Schucknecht2017]. The main environmental variables that affect productivity are water supply (i.e., precipitation) and demand (i.e., evapotranspiration)[@Mishra2015] commonly collected from weather stations. In developing countries, such as Chile, gaps in historical records are challenging. However,  satellite products can be used to fill these gaps.

To evaluate drought, the World Meteorological Organization (WMO; @WMO2012) has proposed the Standardized Precipitation Index (SPI; @Mckee1993), a multiscalar drought index that allows to monitor precipitation deficit from short- to long-term. For water supply, @Zambrano2017 has already correlated i tInfraRed Precipitation against weather  Station data (CHIRPS; @Funk2015). On the other hand, vegetation biomass productivity also strongly correlates to ET, which in turns depends on the atmospheric evaporative demand (AED) [@FAO66; @Pereira2015; @Allen2005]. Due to increasing trends in air temperatures, AED is increasing, driving a rise in ET [@IPCCCH112021]. But, it is not always true [@Milly2016]. For example, regions where AET is highest have the lowest ET.

The MOD16 product [@Running2021] provides AET and ET satellite estimates and has been used to derive drought indices [@Mu2013]. Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates vegetative growth. The climate change initiative (CCI) from the European Space Agency (ESA) delivers the ESA CCI SM product [@Dorigo2017] (current version 6.1), which has been helpful to monitor drought [@Zhang2019]. Besides, total water storage can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes [@Ahmed2014; @Ma2017]. The balance between water demand and supply by remote sensing can be examined against vegetation productivity, also derived from remote sensing.

We aim to analyze the impact of drought over different types of land cover classes in continental Chile for 2000-2023 by combining environmental variables of biomass productivity and water demand/supply gathered from earth observation products. The specific objective for the study are i) to calculate multi-scalar drought indices for water demand and supply for 1981-2023, ii) to analyze the relationship of a proxy o biomass (zcNDVI) with drought indices, iii) to evaluate LULC change for 2001-2021 and its relation with drought indices,  and  iv) to assess if the observed changes in drought indices and land cover are linked to the TWS.

@@

In this study, RL06.1_V3 GRACE mascon (mass concentration) monthly solutions were used, which are provided by the Jet Propulsion Laboratory (JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at 0.5 degrees represents the deviation of surface mass for that month relative to a reference time average (2004-2009 baseline), which is subtracted from all other monthly grids to provide terrestrial water storage anomalies (TWSA) [@Ramjeawon2022; @Yin2022]. This JPL-M version of the data employs a Coastal Resolution Improvement (CRI) filter that reduces signal leakage errors across coastlines [@Wiese2019; @Wiese2016]. Although the mascon solutions greatly reduced leakage errors, a gain-factor, which is used to enhance the spatial resolution, was applied to the dataset as recommended for hydrological studies [@Ramjeawon2022; @Yin2022]. The water storage/height anomalies are given in Equivalent Water Height or Thickness units (EWH, cm) [@Sabzehee2023], and its temporal resolution monthly is from April 2002 to May 2023.

>>>>

title: Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from the ERA5-Land and MODIS

@@

    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'

@@

Precipitation is the primary driver of drought and is intensified by temperature [@Luo2017]. Drought impacts soil moisture, hydrological regimes, and vegetation productivity. Initially, drought was commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans' decisions and activities that drives the anthropogenic drought. Even though it has been argued that those definitions do not fully address the ecological dimensions of drought. @Crausbay2017 proposed the ecological drought definition as *"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"*. Moreover, many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019]. On the other hand, the AR6 [@IPCC2023] states that even if global warming is stabilized at 1.5°–2°C, many parts of the world will be impacted by more severe agricultural and ecological droughts. Then, there is a challenge in conducting drought research, especially to evaluate its impact on ecosystems.

Chile has been facing a persistent rainfall deficit for more than a decade [@Garreaud2017], which has impacted vegetation development [@Zambrano2023] and the hydrological system [@Boisier2018]. Current drought conditions have affected crop productivity [@Zambrano2016; @Zambrano2018], forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], land cover change [@Fuentes2021], water supply in watersheds [@AlvarezGarreton2021], and have had economic impacts [@Fernandez2023]. In 2019–2020, the drought severity reached an extreme condition in Central Chile (30–34°S) not seen for at least 40 years, and the evidence indicates that the impact is transversal to the land cover classes of forest, grassland, and cropland [@Zambrano2023]. The prolonged lack of precipitation in Central Chile is producing changes in ecosystem dynamics that must be studied.

For the spatiotemporal assessment of drought impact (i.e., by water supply and demand) on land cover changes, we need climatic realiable variables such as precipitation, temperature, soil moisture, land cover, and vegetation status. For developing countries like Chile, the weather networks present several disadvantages, such as gaps, a short history, and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L) [@MunozSabater2021] provides hourly climatic information (precipitation, temperature, and soil moisture) without gaps since 1950 with global extension. ERA5L has already been used for drought assessment using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@Nouri2023] and for flash drought [@Wang2023] by analyzing soil mositure and evapotranspiration. On the other hand, satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how drought impacts vegetation dynamics. Vegetation drought indices (VDI) are commonly used as proxies of productivity [@Paruelo2016; @Schucknecht2017], which can be derived from the MODIS (Moderate-Resolution Imaging Spectroradiometer).  Besides, land use and land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021]. To analyze these changes, multiple LULC products exist [@Grekousis2015], one of those that provides time series since 2001 is the MCD12Q1 [@Friedl2019] from MODIS. The variation in water supply and demand is finally reflected in the total water storage (TWS). The TWS can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes at coarse resolution [@Ahmed2014; @Ma2017]. We can use climatic reanalysis (ERA5L) and vegetation data (MODIS) to derive drought indices of supply (i.e., precipitation) and demand (i.e., temperature) and thus evaluate the impact of drought on LULC changes. Further, the TWS can be assessed with regard to the changes in water supply and demand to gain insight into the impact on water storage.

To evaluate meteorological drought (i.e., water supply), the World Meteorological Organization (WMO; @WMO2012) recommends the Standardized Precipitation Index (SPI; @Mckee1993), a multiscalar drought index that allows to monitor precipitation deficits from short- to long-term. Following the same approach, @Vicente-Serrano2010 incorporates into the SPI the effect of temperature through the use of potential evapotranspiration, thus proposing the SPEI (Standardized Precipitation Evapotranspiration Index). Similarly, to evaluate solely the evaporative demand driven by temperature, @Hobbins2016 and @McEvoy2016 came up with the Evaporative Demand Drought Index (EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;  @Zambrano2018 proposed the zcNDVI, a standardized anomaly of the cumulative Normalized Difference Vegetation Index (NDVI), which could be acumulated over the growing season or any period (e.g., months), resulting in a multiscalar drought index. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a normalized index [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021] which is a normalized index using vegetation, land surface temperature, and a vegetation condition index (VCI, [@Kogan1995]). From TWS, we can estimate the standardized terrestrial water storage index (STI) [@Cui2021], a standardized anomaly that follows the methodology of the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for water supply, demand, and storage, which can help to make a comprehensive assessment of drought.

In this research, we aim to analyze the impact of drought on different types of land cover classes in continental Chile by combining environmental variables such as biomass productivity, and water demand and supply gathered from earth observation products. The specific objectives of the study are: i) to calculate multi-scalar drought indices for water demand and supply for 1981–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to the TWS.

@@

In this study, RL06.1_V3 GRACE mascon (mass concentration) monthly solutions were used, which are provided by the Jet Propulsion Laboratory (JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at 0.5 degrees represents the deviation of surface mass for that month relative to a reference time average (2004-2009 baseline), which is subtracted from all other monthly grids to provide terrestrial water storage anomalies (TWSA) [@Ramjeawon2022; @Yin2022]. This JPL-M version of the data employs a Coastal Resolution Improvement (CRI) filter that reduces signal leakage errors across coastlines [@Wiese2019; @Wiese2016]. Although the mascon solutions greatly reduced leakage errors, a gain factor, which is used to enhance the spatial resolution, was applied to the dataset as recommended for hydrological studies [@Ramjeawon2022; @Yin2022]. The water storage and height anomalies are given in Equivalent Water Height or Thickness units (EWH, cm) [@Sabzehee2023], and their temporal resolution monthly is from April 2002 to May 2023.

### 2023-10-09 article/paper_drought_Chile2023.tex
first draft of the introduction

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought assessment of the impact of water supply and demand over land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS datasets},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Drought assessment of the impact of water supply and demand over
land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS
datasets}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl}

@@

\begin{keyword}
    drought \sep land cover change \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Precipitation is the primary driver of drought, which impacts soil
moisture, hydrological regimes, and vegetation productivity. Initially,
drought was commonly classified as meteorological, hydrological, and
agricultural \citep{Wilhite1985}. Lately, \citep{Loon2016} and
\citep{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought. \citep{Crausbay2017}
proposed the ecological drought definition as \emph{``an episodic
deficit in water availability that drives ecosystems beyond thresholds
of vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''}. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} states that even if global warming
is stabilized at 1.5°--2°C, many parts of the world will be impacted by
more severe agricultural and ecological droughts. Then, there is a
challenge in conducting drought research, especially its impact on
ecosystems.

Chile has been facing a persistent rainfall deficit for more than a
decade \citep{Garreaud2017}, which has impacted the hydrological system
\citep{Boisier2018} and consequently the vegetation development
\citep{Zambrano2023}. Current drought conditions have affected crop
yields in Central Chile. Highlighting the growing seasons 2007-2008 and
2008-2009 \citep[\citep{Zambrano2018}]{Zambrano2016}, which impacted an
extensive surface in {[}COMPLETAR{]}. But, in 2019--2020, the drought
intensity reached an extreme condition in North 34°S not seen for at
least 40 years \citep{Zambrano2023}, thus affecting forest, grassland,
and cropland areas. The prolonged lack of precipitation in Central Chile
is producing changes in ecosystem dynamics that must be studied.

Satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how meteorological drought impacts vegetation
dynamics. Since the 90's multiple vegetation drought indices have been
developed, such as (VCI,\citep{Kogan1990}; TCI, \citep{Kogan1995};
zNDVI, \citep{Peters2002}; VegDri, \citep{Brown2008}) that have allowed
making spatiotemporal analysis. Although we can derive these indices for
any time during the year (depending on satellite revisit), they are more
relevant during the the growing season \citep{Mishra2015}. Although
modeling phenology is a complex task (e.g, \ldots.), satellites offer
strategies that help to address these issues
\citep{Younes2021, Vrieling2018, Cai2017}. Moreover, land cover dynamics
products such as the MCD12Q2 from the USGS \citep{Friedl2019} provide
phenology metrics. Rather than estimating these parameters at a given
point in time, some authors have proposed aggregating these indices
throughout the season. \citep{Meroni2017} proposes to accumulate the
fractional active photosynthetic active radiation (FAPAR) between the
start (SOS), and the end of the season (EOS) thus obtaining zCFAPAR as a
proxy for biomass productivity. \citep{Zambrano2018} used the same
approach but with the NDVI (Normalized Difference Vegetation Index),
deriving the zcNDVI for Central Chile. Besides, land use land cover
(LULC) change can be driven by drought \citep{Tran2019, Akinyemi2021}.
To analyze these changes, multiple time-series LULC products exist, such
as the MCD12Q1 \citep{Friedl2019} and the ESA CCI-LC (ESA 2017). The
LULC product, together with the vegetation drought index can help
evaluate the impact of drought on the ecosystem.

Vegetation drought indices are commonly used as proxies of productivity
\citep{Paruelo2016, Schucknecht2017}. The main environmental variables
that affect productivity are water supply (i.e., precipitation) and
demand (i.e., evapotranspiration)\citep{Mishra2015} commonly collected
from weather stations. In developing countries, such as Chile, gaps in
historical records are challenging. However, satellite products can be
used to fill these gaps.

To evaluate drought, the World Meteorological Organization (WMO;
\citep{WMO2012}) has proposed the Standardized Precipitation Index (SPI;
\citep{Mckee1993}), a multiscalar drought index that allows to monitor
precipitation deficit from short- to long-term. For water supply,
\citep{Zambrano2017} has already correlated i tInfraRed Precipitation
against weather Station data (CHIRPS; \citep{Funk2015}). On the other
hand, vegetation biomass productivity also strongly correlates to ET,
which in turns depends on the atmospheric evaporative demand (AED)
\citep{FAO66, Pereira2015, Allen2005}. Due to increasing trends in air
temperatures, AED is increasing, driving a rise in ET
\citep{IPCCCH112021}. But, it is not always true \citep{Milly2016}. For
example, regions where AET is highest have the lowest ET.

The MOD16 product \citep{Running2021} provides AET and ET satellite
estimates and has been used to derive drought indices \citep{Mu2013}.
Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates
vegetative growth. The climate change initiative (CCI) from the European
Space Agency (ESA) delivers the ESA CCI SM product \citep{Dorigo2017}
(current version 6.1), which has been helpful to monitor drought
\citep{Zhang2019}. Besides, total water storage can be retrieved by the
Gravity Recovery and Climate Experiment (GRACE), which allows analyzing
water availability changes \citep{Ahmed2014, Ma2017}. The balance
between water demand and supply by remote sensing can be examined
against vegetation productivity, also derived from remote sensing.

We aim to analyze the impact of drought over different types of land
cover classes in continental Chile for 2000-2023 by combining
environmental variables of biomass productivity and water demand/supply
gathered from earth observation products. The specific objective for the
study are i) to calculate multi-scalar drought indices for water demand
and supply for 1981-2023, ii) to analyze the relationship of a proxy o
biomass (zcNDVI) with drought indices, iii) to evaluate LULC change for
2001-2021 and its relation with drought indices, and iv) to assess if
the observed changes in drought indices and land cover are linked to the
TWS.

@@

In this study, RL06.1\_V3 GRACE mascon (mass concentration) monthly
solutions were used, which are provided by the Jet Propulsion Laboratory
(JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at
0.5 degrees represents the deviation of surface mass for that month
relative to a reference time average (2004-2009 baseline), which is
subtracted from all other monthly grids to provide terrestrial water
storage anomalies (TWSA) \citep{Ramjeawon2022, Yin2022}. This JPL-M
version of the data employs a Coastal Resolution Improvement (CRI)
filter that reduces signal leakage errors across coastlines
\citep{Wiese2019, Wiese2016}. Although the mascon solutions greatly
reduced leakage errors, a gain-factor, which is used to enhance the
spatial resolution, was applied to the dataset as recommended for
hydrological studies \citep{Ramjeawon2022, Yin2022}. The water
storage/height anomalies are given in Equivalent Water Height or
Thickness units (EWH, cm) \citep{Sabzehee2023}, and its temporal
resolution monthly is from April 2002 to May 2023.

@@

>>>>

\usepackage{lineno}\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from the ERA5-Land and MODIS},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Assessment of drought impact over land cover in continental Chile
by the analysis of water supply and demand from the ERA5-Land and MODIS}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl}

@@

\begin{keyword}
    drought \sep land cover change \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

Precipitation is the primary driver of drought and is intensified by
temperature \citep{Luo2017}. Drought impacts soil moisture, hydrological
regimes, and vegetation productivity. Initially, drought was commonly
classified as meteorological, hydrological, and agricultural
\citep{Wilhite1985}. Lately, \citep{Loon2016} and
\citep{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought. \citep{Crausbay2017}
proposed the ecological drought definition as \emph{``an episodic
deficit in water availability that drives ecosystems beyond thresholds
of vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''}. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} states that even if global warming
is stabilized at 1.5°--2°C, many parts of the world will be impacted by
more severe agricultural and ecological droughts. Then, there is a
challenge in conducting drought research, especially to evaluate its
impact on ecosystems.

Chile has been facing a persistent rainfall deficit for more than a
decade \citep{Garreaud2017}, which has impacted vegetation development
\citep{Zambrano2023} and the hydrological system \citep{Boisier2018}.
Current drought conditions have affected crop productivity
\citep{Zambrano2016, Zambrano2018}, forest development
\citep{Miranda2020, Venegas2018}, forest fire occurrence
\citep{UrrutiaJalabert2018}, land cover change \citep{Fuentes2021},
water supply in watersheds \citep{AlvarezGarreton2021}, and have had
economic impacts \citep{Fernandez2023}. In 2019--2020, the drought
severity reached an extreme condition in Central Chile (30--34°S) not
seen for at least 40 years, and the evidence indicates that the impact
is transversal to the land cover classes of forest, grassland, and
cropland \citep{Zambrano2023}. The prolonged lack of precipitation in
Central Chile is producing changes in ecosystem dynamics that must be
studied.

For the spatiotemporal assessment of drought impact (i.e., by water
supply and demand) on land cover changes, we need climatic realiable
variables such as precipitation, temperature, soil moisture, land cover,
and vegetation status. For developing countries like Chile, the weather
networks present several disadvantages, such as gaps, a short history,
and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L)
\citep{MunozSabater2021} provides hourly climatic information
(precipitation, temperature, and soil moisture) without gaps since 1950
with global extension. ERA5L has already been used for drought
assessment using the Standardized Precipitation-Evapotranspiration Index
(SPEI) \citep{Nouri2023} and for flash drought \citep{Wang2023} by
analyzing soil mositure and evapotranspiration. On the other hand,
satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how drought impacts vegetation dynamics.
Vegetation drought indices (VDI) are commonly used as proxies of
productivity \citep{Paruelo2016, Schucknecht2017}, which can be derived
from the MODIS (Moderate-Resolution Imaging Spectroradiometer). Besides,
land use and land cover (LULC) change can be driven by drought
\citep{Tran2019, Akinyemi2021}. To analyze these changes, multiple LULC
products exist \citep{Grekousis2015}, one of those that provides time
series since 2001 is the MCD12Q1 \citep{Friedl2019} from MODIS. The
variation in water supply and demand is finally reflected in the total
water storage (TWS). The TWS can be retrieved by the Gravity Recovery
and Climate Experiment (GRACE), which allows analyzing water
availability changes at coarse resolution \citep{Ahmed2014, Ma2017}. We
can use climatic reanalysis (ERA5L) and vegetation data (MODIS) to
derive drought indices of supply (i.e., precipitation) and demand (i.e.,
temperature) and thus evaluate the impact of drought on LULC changes.
Further, the TWS can be assessed with regard to the changes in water
supply and demand to gain insight into the impact on water storage.

To evaluate meteorological drought (i.e., water supply), the World
Meteorological Organization (WMO; \citep{WMO2012}) recommends the
Standardized Precipitation Index (SPI; \citep{Mckee1993}), a multiscalar
drought index that allows to monitor precipitation deficits from short-
to long-term. Following the same approach, \citep{Vicente-Serrano2010}
incorporates into the SPI the effect of temperature through the use of
potential evapotranspiration, thus proposing the SPEI (Standardized
Precipitation Evapotranspiration Index). Similarly, to evaluate solely
the evaporative demand driven by temperature, \citep{Hobbins2016} and
\citep{McEvoy2016} came up with the Evaporative Demand Drought Index
(EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
~\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
cumulative Normalized Difference Vegetation Index (NDVI), which could be
acumulated over the growing season or any period (e.g., months),
resulting in a multiscalar drought index. For soil moisture, several
drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a
normalized index \citep{Narasimhan2005} and the Soil Moisture
Agricultural Drought Index (SMADI) \citep{Souza2021} which is a
normalized index using vegetation, land surface temperature, and a
vegetation condition index (VCI, \citep{Kogan1995}). From TWS, we can
estimate the standardized terrestrial water storage index (STI)
\citep{Cui2021}, a standardized anomaly that follows the methodology of
the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for
water supply, demand, and storage, which can help to make a
comprehensive assessment of drought.

In this research, we aim to analyze the impact of drought on different
types of land cover classes in continental Chile by combining
environmental variables such as biomass productivity, and water demand
and supply gathered from earth observation products. The specific
objectives of the study are: i) to calculate multi-scalar drought
indices for water demand and supply for 1981--2023; ii) to evaluate LULC
change for 2001--2021 and its relation to drought indices; iii) to
analyze the relationship of a proxy of biomass (zcNDVI) with drought
indices; and iv) to assess if the observed changes in the drought
indices are linked to the TWS.

@@

In this study, RL06.1\_V3 GRACE mascon (mass concentration) monthly
solutions were used, which are provided by the Jet Propulsion Laboratory
(JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at
0.5 degrees represents the deviation of surface mass for that month
relative to a reference time average (2004-2009 baseline), which is
subtracted from all other monthly grids to provide terrestrial water
storage anomalies (TWSA) \citep{Ramjeawon2022, Yin2022}. This JPL-M
version of the data employs a Coastal Resolution Improvement (CRI)
filter that reduces signal leakage errors across coastlines
\citep{Wiese2019, Wiese2016}. Although the mascon solutions greatly
reduced leakage errors, a gain factor, which is used to enhance the
spatial resolution, was applied to the dataset as recommended for
hydrological studies \citep{Ramjeawon2022, Yin2022}. The water storage
and height anomalies are given in Equivalent Water Height or Thickness
units (EWH, cm) \citep{Sabzehee2023}, and their temporal resolution
monthly is from April 2002 to May 2023.

@@

\hypertarget{drought-indices-for-water-demand-and-supply-1}{%
\subsection{Drought indices for water demand and
supply}\label{drought-indices-for-water-demand-and-supply-1}}
